Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2007-07-07 SCP2008/paper/metamonads.tex
editing and rearranging of the final section.

Parameterizing by the kind of container representing a matrix is
almost as straightforward.  The type of our containers include the
lifted domain |Dom| as one of the component. Particular containers are
Our containers are  parametric
over a |DOMAINL|, i.e., functors from a |DOMAINL| module
to the actual implementation of a container. For example, the
following functor defines a matrix container represented as arrays of
rows.

\begin{code}
module GenericArrayContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x n m = .< (.~x).(.~n).(.~m) >.
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  \dots
\end{code}
%
We also have containers wheer elements are stored in a 1D array, in a
row-wise (C-like) or column-wise (Fortran-like) modes.

The use of a |functor| for making a container parametric is fairly
straightforward.  More interesting is the aspect of what to return
from the LU algorithm.  One could create an algebraic data type (as
was done in \cite{carette04}) to encode the various choices: the
matrix, the matrix and the rank, the matrix and the determinant, the
matrix, rank and determinant, and so on. This is wholly unsatisfying
as we know that for any single use, only one of the choices is ever
possible, yet any routine which calls the generated code must deal
with these unreachable options.  Instead we use a module type with an
\emph{abstract} type |res| for the result type; different instances of
the signature set the result type differently. Given below is this
module type and one instantiation, which specifies the output of a LU
algorithm as a 3-tuple |contr * Det.outdet * int| of the U-factor, the
determinant and the rank.

@@

\subsection{Maintaining the state}
another title: representing extensible state

different aspects may need to keep their own state. We wish to assure
modularity. One approach: using objects. The state is an object, each
aspect is an instnace variable with some particular name. The problem
is how to represent the initial `empty' state. Need to create the
object with all fields set to None. But we need to know the names of
all fields; as we add an aspect, our function that runs the generator
must change to account for the different initial object with different
slots.

A better approach: property list (ref MLTon). Property list represents
an extensible object. The initial object is just the empty list. Alas,
the literal MLton's approach doesn't apply to us well. First of all,
due to the environment classifiers, we need universes parameterized by
classifiers. Also, the propoerty names are generated automatically in
MLton approach, based on generativity of exceptions or reference
cells. That means if some particular aspect happens to be included
twice, the state will be incompatible. We will have to worry about
sharing or the lack of sharing of our aspects. We would rather prefer
our aspects to be stateless and work the same way whether two
instances of the aspect are shared or not. So, we find polymorphic
unions to be ideal. Manifest naming

Encoding an object by its dual,
list of polymorphic variants.

Currently, we check at genertor-time that one cannot add a slot which
already exists (already the part of the state) nor we can obtain the
value of a non-existing slot. These problems may occur if our genetor
is wrong (calls fin method before decl method of an aspect). It is
possible to make these checks static (TFP2007, state-variable monad?)
state-changing monad...

\subsection{tracking}

As is apparent from the output choices, several different quantities
\emph{may} need to be tracked in a particular LU implementation.  We
therefore need to be able to conditionally generate variables
representing the tracking state, and weave in corresponding tracking
code. We may need to (independently) keep track of the rank, the
determinant and the permutation list.  The tracking state variables
then become part of the \emph{state} that is tracked by our monad.  To
have all this choice when needed, and yet have our code be modular and
composable as well as ensuring that the generated code does not
contain any abstraction artifacts, it is important to make this state
modular.  For example,
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val get       : unit -> ('b,tdet) lm
  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
\noindent  to track determinant we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign or the absolute
value of the determinant, converting the tracking state
to the final determinant value of the type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant, even if we don't output
it. The determinant tracking aspect is complex because tracking
variables, if any, are to be declared at the beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. Other aspects, e.g., rank tracking, may use the
monadic state for passing of rank tracking variables. To be able to
compose determinant and rank tracking functors -- each of which may
(or may not) use the monadic state for passing its own data -- we make
extensive use of open records (a list of polymorphic variants
appeared to be the easiest way to implement such a union, in a purely
functional way). This lets us freely compose determinant-tracking,
rank-tracking, and other aspects.

We have several instances of Det
\begin{code}
module NoDet =
  struct
  type tdet = C.Dom.v ref
  type 'a lstate = unit
  let decl () = unitL
  let upd_sign () = ret None
  let zero_sign () = unitL
  let acc _ = unitL
  let get () = ret (liftRef C.Dom.zeroL) (* hack alert! *)
  let set _ = unitL
  let fin () = failwith "Determinant is needed but not computed"
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
end

module AbstractDet =
  struct
  open C.Dom
  type tdet = v ref
  (* the first part of the state is an integer: which is +1, 0, -1:
     the sign of the determinant *)
  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
  let fin = fun () -> perform
      (det_sign,det) <-- mo_lookup ip;
      ifM (Logic.equalL (liftGet det_sign) Idx.zero) (ret zeroL)
      (ifM (Logic.equalL (liftGet det_sign) Idx.one) (ret (liftGet det))
          (ret (uminusL (liftGet det))))
end
\end{code}

Both contain the |fin ()| function to generate the code representing
the computed rank. Only the |NoRank| does not track any rank, and so 
|fin ()| raises an error. The code 
|let _ = IF.R.fin ()| in |OutDetRank| invokes that fin function, which
will produce the monadic code generating action, or raise teh
error. We do not run the action at that time -- we only make sure
there is an action to run. This is the preflight check again, to rule
out the semantic error when the user specified that rank should be
computed and returned and yet specified the |NoRank| aspects which
computes no rank at all.

\begin{code}
(* The `keyword' list of all the present external features *)
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

@@

The LU generator functor itself is 
parameterized by the domain, container, pivoting policy (full, row,
nonzero, no pivoting), update policy (with either `fraction-less'
or full division), and the result specification. Some of the
argument modules such as |PIVOT| are functors themselves (parameterized
by the domain, the container, and the determinant functor). The sharing
constraints express obvious constraints on the instantiation of |Gen|,
for example, pivoting, determinant etc. components all use the same
domain. It must be stressed that all structures (i.e., module
instances) are stateless, and so we never have to worry that different
aspect functors (such as |CONTAINER2D| and |PIVOT|) are instantiated
with different but type-compatible instances of |DOMAIN|. That is, we
are not concerned at all about value sharing. Aspects such as
determinant tracking may be stateful so that the determinant update
code have access to the determinant tracking variables declared
previously. But that state is handled via the monadic state. As we
have shown, open unions make the overall monadic state compositional
with respect to the state of various aspects.

In addition to the ``regular'' type sharing constraints shown in the
|Gen| functor, there are also ``semantic'' sharing constraints, shown
in the following structure of the |UPDATE| signature:
\vspace*{-5pt}\begin{code}
module DivisionUpdate
  (Dom:DOMAIN with type kind = domain_is_field)
  (C:CONTAINER2D)
  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
\end{code}
\vspace*{-4pt} 
This structure implements an update policy of using
|Dom.div| operation without restrictions -- which is possible only if
the domain has such an unrestricted operation. A domain such as the integer
domain may still provide |Dom.div| of the same type, but that operation may
only be used when we are sure that the division is exact. Our type
sharing constraint expresses such domain-specific knowledge:
instantiating |DivisionUpdate| with |IntegerDomain| leads to a
compile-time error, when compiling the \emph{generator} code. Thus, in
some cases we can use module types for ``semantic'' constraints
that cannot normally be expressed via the types of module members.
\vspace*{-5pt}
\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)

module GAC_F = GenericArrayContainer(FloatDomainL)
module G_GAC_F = GenLA(GAC_F)
open G_GAC_F
open G_GAC_F.GE

module GenFA1 = GenGE(struct
    module Det = NoDet
    module PivotF = RowPivot
    module PivotRep = PermList
    module Update = DivisionUpdate
    module Input = InpJustMatrix 
    module Output = OutJustMatrix end)
module GenFA9 = GenGE(struct 
    module Det = NoDet
    module PivotF = RowPivot
    module PivotRep = PermList
    module Update = DivisionUpdate
    module Input = InpJustMatrix
    module Output = Out_LU_Packed end)

let resIV5 = instantiate GenIV5.gen ;;
let rIV5 = runit {pf =  resIV5 };;

\end{code}
|GenFA1| is no-frills GE; GenFA9 has LU part

We can instantiate the |Gen| functor as shown above and inspect the generated
code, e.g., by printing |GenFA1.gen|. The code can then be ``compiled'' as 
|!. GenFA1.gen| or with off-shoring. The code for |GenIV5| (Appendix A) shows
full pivoting, determinant and rank tracking. The code for all these aspects is
fully inlined; no extra functions are invoked and no tests other than those
needed by the LU algorithm itself are performed. The GE?? function returns a
triple |int array * int * int| of the U-factor, determinant and the rank. The
code generated by |GenFA1| (Appendix B) shows absolutely no traces of
determinant tracking: no declaration of spurious variables, no extra tests,
etc. The code appears as if the determinant tracking aspect did not exist
at all. The generated code for the above and other instantiations of
|Gen| can be examined at \cite{metamonadsURL}. The website also 
contains benchmark code and timing comparisons.

@@

The code generated for |GenFA9|, LU of the floating point matrix
represented by a 2D array, row pivoting, returning just the |U|-factor.

>>>>

The type of our containers include the lifted domain |Dom| as one of
the component. This is quite convenient since operations on containers
are usually accompanied by the operations on the retrieved and stored
values. Our containers are parametric over a |DOMAINL|, i.e., functors
from a |DOMAINL| module to the actual implementation of a
container. For example, the following functor defines a matrix
container represented as arrays of rows.

\begin{code}
module GenericArrayContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x n m = .< (.~x).(.~n).(.~m) >.
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  \dots
\end{code}
%
The accompanying code includes containers whose
elements are stored in a 1D array, in a
row-wise (C-like) or column-wise (Fortran-like) modes.

\subsection{Maintaining the extensible state}

Various aspects of our LU generator such as determinant
computation, rank, permutation list may need to keep a state during
the code generation. For example, the initializing section of the LU
generator will invoke the |decl| of the determinant aspect so that the
latter may generate let-binding for reference cells that track the
magnitude and the sign of the determinant. The generator of the
row-swapping code will notify the determinant aspect so that the
latter may insert the code that flips the sign of the tracked
determinant. Finally, the final section of the generator will ask the
determinant aspect to generator code producing the final, signed value
of the determinant. Clearly the determinant aspect needs to maintain
state from one invocation of its methods to another; this state should
include the `names' of mutable variables accumulating the sign and the
magnitude of the determinant.

The simplest method of keeping the state of the generator is by using
mutable variables private to the Deteminant module (aspect). That,
however, makes our aspects stateful. Although we are generating
imperative code, we would like to keep our generators stateless and
pure functional, for ease of comprehension and reasoning. Also, if a
module is stateless, we do not have to worry about sharing, aliasing
and multiple instances. Our main program may include several LU
generators, which may refer to the Determinant aspect -- which may be
present in one shared instance or in several. It is irrelevant for us
which is which. With stateful modules, the issues of aliasing or
separate instantiation are the source of very subtle problems.

We chose therefore a different way of maintaining the generator state,
using a monad. We need monad anyway for let-insertion; we add state as
an additional monad parameter (Fig.~\ref{our-monad}) and monadic
actions |fetch| and |store| to access that monadic state. The state is
threaded throughout the entire code-generation computation.

This monadic state has to accommodate several pieces of state, for
various aspects. We should be able to add a new aspect, which may need
to keep its own state as part of teh overall monadic state, without
modifying or even recompiling the rest of the code. Thus our monadic
state should be extensible. One way of achieving is via an extensible
record, or OCaml object. Each aspect will have its own field; record
subtyping will assure modularity. Alas, this approach makes it
difficult to create the initial state, to pass to the monad's |runM|
method. We should know the names of all fields and should know the
proper initial value for these fields. This breaks the modularity.

A better approach of property list has been suggested my the MLTon
team \oleg{cite MLTon}.  Encoding an object by its dual, list of
polymorphic variants.  Property list represents an extensible
object. The initial object is just the empty list. Alas, the MLton's
approach does not literally apply to us. First of all, due to
the environment classifiers, we need universes parameterized by
classifiers. Also, the property names are generated automatically
in MLton approach, based on generativity of exceptions or reference
cells. That means if some particular aspect happens to be included
twice, the state will be incompatible. We will have to worry about
sharing or the lack of sharing of our aspects. We would rather
prefer our aspects to be stateless and work the same way whether two
instances of the aspect are shared or not. Fortunately, OCaml offers a
way to build open unions using polymorphic variants. Each variant is
identified by a manifest tag, e.g., `Tdet, and may include a
value. The tags are not generative and provide the form of manifest
naming (somewhat similar to symbols of Lisp and Scheme).

Below is the implementation of our open records with manifest
naming, functions |orec_store| to add a new field to the open record
and |orec_find| to obtain the value associated with a particular field
name. Each `field' is characterized by a triple: injection, 
projection functions and the string name. The latter is used for 
printing error messages. For example, for the
determinant tracking aspect, this triple has the form
\begin{code}
  let ip = (fun x -> `TDet x), 
           (function `TDet x -> Some x | _ -> None), 
           "Det"
\end{code}
We combine these functions with monadic
actions to access monadic state and so obtain 
|mo_extend| and |mo_lookup| to store and retrieve one particular value
of the monadic state.
\begin{code}
let rec lookup ((_,prj,_) as ip) = 
   function [] -> raise Not_found
   | (h::t) -> (match prj h with Some x -> x | _ -> lookup ip t)

let orec_store ((inj,_,name) as ip) v s =
  let () = 
    try let _ = lookup ip s in 
        failwith ("The field of an open record is already present: " ^ name)
    with Not_found -> () in
  (inj v)::s

let orec_find ((_,_,name) as ip) s =
  try lookup ip s 
  with Not_found -> failwith ("Failed to locate orec field: " ^ name)

let mo_extend ip v = 
  perform s <-- fetch; store (orec_store ip v s)

let mo_lookup ip =
  perform s <-- fetch; ret (orec_find ip s)
\end{code}

Currently, we check at generator-time that one cannot add a slot which
already exists (already the part of the state) nor we can obtain the
value of a non-existing slot. These problems may occur if our genetor
is wrong (calls fin method before decl method of an aspect). It is
possible to make these checks static (TFP2007, state-variable monad?)
state-changing monad...

\subsection{Determinant aspect}

To track determinant we should be able to generate code
for: defining variables used for tracking (|decl|),
updating the sign or the absolute
value of the determinant, converting the tracking state
to the final determinant value of the type |outdet|. LU of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant, even if we don't output
it. The determinant tracking aspect is complex because tracking
variables, if any, are to be declared at the beginning of LU; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, e.g., a piece of
code for the value of the type |Dom.v ref|, among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state.

Other aspects follow a similar outline

\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  val decl : unit -> ('b,unit) lm (* could be unit rather than unit code...*)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val get       : unit -> ('b,tdet) lm
  val set       : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}

We have two instances of |Det|. The first one is used when the user
requested that no determinant tracking is needed, and so no code is
generated. The methods just return without executing any
code-generation action.
\begin{code}
module NoDet =
  struct
  type tdet = C.Dom.v ref
  type 'a lstate = unit
  let decl () = unitL
  let upd_sign () = ret None
  let zero_sign () = unitL
  let acc _ = unitL
  let get () = ret (liftRef C.Dom.zeroL)
  let set _ = unitL
  let fin () = failwith "Determinant is needed but not computed"
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
end
\end{code}

The following instance does track the determinant. The |decl| method
generates two let-bindings for mutable variables tracking the
magnitude and the sign of the determinant, and places the names of
these variables in the monadic state. The |upd_sign| method, invoked
from row- or column-swap generators, retrieves the name of the
sign-accumulating variable and generates the code to update the
sign. The |fin| method retrieves the names of both accumulators and
generates code to compute the final, signed determinant value.
\begin{code}
module AbstractDet =
  struct
  open C.Dom
  type tdet = v ref
  (* the first part of the state is an integer: which is +1, 0, -1:
     the sign of the determinant *)
  type 'a lstate = ('a,int ref) abstract * ('a,tdet) abstract
  let decl () = perform
      ddecl <-- retN (liftRef oneL);
      dsdecl <-- retN (liftRef Idx.one);
      mo_extend ip (dsdecl,ddecl);
      unitL
  let upd_sign () = perform
      det <-- mo_lookup ip;
      det1 <-- ret (fst det);
      ret (Some (assign det1 (Idx.uminus (liftGet det1))))
  let fin = fun () -> perform
      (det_sign,det) <-- mo_lookup ip;
      ifM (Logic.equalL (liftGet det_sign) Idx.zero) (ret zeroL)
      (ifM (Logic.equalL (liftGet det_sign) Idx.one) (ret (liftGet det))
          (ret (uminusL (liftGet det))))
  \dots
end
\end{code}

\subsection{Output}

More interesting is the aspect of what to return
from the LU algorithm.  One could create an algebraic data type (as
was done in \cite{carette04}) to encode the various choices: the
matrix, the matrix and the rank, the matrix and the determinant, the
matrix, rank and determinant, and so on. This is wholly unsatisfying
as we know that for any single use, only one of the choices is ever
possible, yet any routine which calls the generated code must deal
with these unreachable options.  Instead we use a module type with an
\emph{abstract} type |res| for the result type; different instances of
the signature set the result type differently. Given below is this
module type and one instantiation, which specifies the output of a LU
algorithm as a 3-tuple |contr * Det.outdet * int| of the U-factor, the
determinant and the rank.

@@

Both instances of |Det| contain the |fin ()| function to generate the
code representing the computed determinant. Only the |NoDet| does not
track any determinant, and so |fin ()| raises an error. The code 
|let _ = OD.Det.fin ()| in |OutDetRank| invokes that |fin| function, which will
produce the monadic code generating action, or raise the error. We do
not run the action at that time -- we only make sure there is an
action to run. This is the preflight check again, to rule out the
semantic error when the user specified that determinant should be computed
and returned and yet specified the |NoDet| aspect which tracks no
determinant at all.

\subsection{Main generation}
We combine all aspects in a record. This record is like a keyword
argument list to the |GenGE| functor below.
\begin{code}
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

\oleg{describe how generation proceeds}
We first instantiate the container |GVC_I| for the vector over an
integer domain. We pass the container to the main functor |GenLA|,
since all the linear algebra aspects are parameterized by the
container and the domain included in it. We open the |G_GVC_I| module
to make all its components easily available. One of thos components is
the GE generrator, |GE| (which we open as well). The module |G_GVC_I|
also inlcudes the generator for GE solvers. The module |GE| includes
the functor |GenGE| that is the generator. We instantiate the
generator by passing the functor the `record' of various aspects (the
order is irrelevant, but all aspects like |Det|, etc. must be
specified). In the code below, we request generation with full pivot,
fraction-free update, operate on non-augmented matrix and return the
|U| factor, determinant and the rank. We run the monad passing the
initial state |[]| and obtain the code in App A.

\begin{code}
module GVC_I = GenericVectorContainer(IntegerDomainL)
module G_GVC_I = GenLA(GVC_I)
open G_GVC_I
open G_GVC_I.GE
module GenIV5 = GenGE(struct 
    module Det = AbstractDet
    module PivotF = FullPivot
    module PivotRep = PermList
    module Update = FractionFreeUpdate
    module Input = InpJustMatrix
    module Output = OutDetRank end)
let instantiate gen =
    .<fun a -> .~(runM (gen .<a>.) []) >.;;
let resIV5 = instantiate GenIV5.gen ;;
let rIV5 = .! resIV5;;

module GAC_F = GenericArrayContainer(FloatDomainL)
module G_GAC_F = GenLA(GAC_F)
open G_GAC_F
open G_GAC_F.GE
module GenFA9 = GenGE(struct 
    module Det = NoDet
    module PivotF = RowPivot
    module PivotRep = PermList
    module Update = DivisionUpdate
    module Input = InpJustMatrix
    module Output = Out_LU_Packed end)
\end{code}

@@

The LU generator functor itself is 
parameterized by the domain, container, pivoting policy (full, row,
nonzero, no pivoting), update policy (with either `fraction-less'
or full division), and the result specification. Some of the
argument modules such as |PIVOT| are functors themselves (parameterized
by the domain, the container, and the determinant functor).

There are more pre-flight checks for various ``semantic'' 
constraints, shown
in the following structure of the |UPDATE| signature:
\vspace*{-5pt}\begin{code}
module DivisionUpdate(Det:DETERMINANT) =
  struct
  open C.Dom = struct 
  let _ = assert (C.Dom.kind = Domains_sig.Domain_is_Field)
end
\end{code}
%
This structure implements an update policy of using
|Dom.div| operation without restrictions -- which is possible only if
the domain has such an unrestricted operation. A domain such as the integer
domain may still provide |Dom.div| of the same type, but that operation may
only be used when we are sure that the division is exact.

We can instantiate the |Gen| functor as shown above and inspect the generated
code, e.g., by printing |resIV5|. The code can then be ``compiled'' as 
|!. resIV5| or with off-shoring. The code for |resIV5| (Appendix A) shows
full pivoting, determinant and rank tracking. The code for all these aspects is
fully inlined; no extra functions are invoked and no tests other than those
needed by the LU algorithm itself are performed. The GE?? function returns a
triple |int array * int * int| of the U-factor, determinant and the rank. The
code generated by |GenFA9| (Appendix B) shows absolutely no traces of
determinant tracking: no declaration of spurious variables, no extra tests,
etc. The code appears as if the determinant tracking aspect did not exist
at all. The generated code for the above and other instantiations of
|Gen| can be examined at \cite{metamonadsURL}. The website also 
contains benchmark code and timing comparisons.

@@

The code generated for |GenFA9|, LU of the floating point
non-augmented matrix
represented by a 2D array, row pivoting, returning the complete
factorization: |L| and |U|
factors packed in a single matrix and the permutation matrix
represented as the list of row number exchanges.
